Longitudinal Principal Manifold Estimation

### 2024-02-12 paper/main.tex
add algorithm udpates

\begin{algorithm}
\caption{Longitudinal Principal Manifold Estimation}\label{alg:lpme}
  \KwData{Data points $\left\{Y_{i, t}\right\}_{i=1, t=1}^{I_t, T}$, positive integer $d$, positive integer $N_0 < \min{I_t} - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\left\{\lambda_k\right\}_{k=1}^K$, $\left\{\gamma_l\right\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
\KwResult{Analytic formula of $\hat{f}^*: \mathbb{R}^{d + 1} \to \mathbb{R}^{D + 1}$, optimal tuning parameter $\gamma^*$.}
\For{$t = 1, 2, \dots, T$} {
  Apply HDMDE algorithm with input $\left(\left\{Y_{i, t}\right\}_{i = 1}^{I_t}, N_0, \epsilon, \alpha\right)$ and obtain $N_t$, $\left\{\mu_{j, t}\right\}_{j = 1}^{N_t}$, and $\left\{\theta_{j, t}\right\}_{j = 1}^{N_t}$\;
}

@@

\For{t = 1, 2, \dots, T} {
  Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^{N_t}$ and obtain $f_t$, $\lambda_t^*$, and $\tau_t$\;
  $\left\{r_{j, t}\right\} \gets \pi_{f_t}(\mu_{j, t})$ for $j = 1, \dots, N_t$ and $X_{j, t} \gets f_t(r_{j, t})$ for $j = 1, \dots, N_t$\;
}

@@

\For{t = 1, 2, \dots, T} {
  Set $X_{i, t} = f_t(r_{i}^*)$ for $i = 1, \dots, N$\;
  Compute $f_t^*(r)$ by solving \eqref{eq:13}\;
  Set $w_t = \frac{\frac{1}{\tau_t}}{\sum_{i=1}^{T}\frac{1}{\tau_i}}$\;
}

@@

A factorial design is used to run the simulation studies, with the factor levels set as follows: 1) $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$, 2) study duration: $\left\{1, 2, 5\right\}$, 3) interval between images: $\left\{0.1, 0.25, 0.5\right\}$, 4) longitudinal change model: Constant, Linear, Quadratic, Sinusoidal. The sample size for each time point is set to 1,000 observations, so each embedding map is simulated a total of 7,776 times. All code required to reproduce the simulation results and the associated figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.

@@

The mean, median, and standard deviation of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Tables \ref{table:simulation_results_mean} and \ref{table:simulation_results_median}. Because the PME, principal curve, and principal surface methods each attempt to estimate the manifold in question at each individual time point without allowing other time points to inform these estimates, they should result in similar deviations from the true underlying manifold, with differences in error resulting primarily from differing performances in fitting to the observed data. Meanwhile, because the LPME approach accounts for all time points simultaneously, this approach should result in lower mean squared distance values.

>>>>

\begin{algorithm}
\caption{Longitudinal Principal Manifold Estimation}\label{alg:lpme}
  \KwData{Data points $\left\{X_{i, t}\right\}_{i=1, t=1}^{I_t, T}$, positive integer $d$, positive integer $N_0 < \min{I_t} - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\left\{\lambda_k\right\}_{k=1}^K$, $\left\{\gamma_l\right\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
\KwResult{Analytic formula of $\hat{f}^*: \mathbb{R}^{d + 1} \to \mathbb{R}^{D + 1}$, optimal tuning parameter $\gamma^*$.}
\For{$t = 1, 2, \dots, T$} {
  Apply HDMDE algorithm with input $\left(\left\{X_{i, t}\right\}_{i = 1}^{I_t}, N_0, \epsilon, \alpha\right)$ and obtain $N_t$, $\left\{\mu_{j, t}\right\}_{j = 1}^{N_t}$, and $\left\{\theta_{j, t}\right\}_{j = 1}^{N_t}$\;
}

@@

\For{t = 1, 2, \dots, T} {
  Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^{N_t}$ and obtain $f_t$, $\lambda_t^*$, and $\tau_t$\;
  $\left\{r_{j, t}\right\} \gets \pi_{f_t}(\mu_{j, t})$ for $j = 1, \dots, N_t$ and $Y_{j, t} \gets f_t(r_{j, t})$ for $j = 1, \dots, N_t$\;
}

@@

\For{t = 1, 2, \dots, T} {
  Set $Y_{i, t} = f_t(r_{i}^*)$ for $i = 1, \dots, N$\;
  Compute $f_t^*(r)$ by solving \eqref{eq:13}\;
  Set $w_t = \frac{1}{\tau_t \sum_{i=1}^{T}\frac{1}{\tau_i}}$\;
}

@@

A factorial design is used to run the simulation studies, with the factor levels set as follows: 1) $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$; 2) study duration: $\left\{1, 2, 5\right\}$; 3) interval between images: $\left\{0.1, 0.25, 0.5\right\}$; 4) longitudinal change model: Constant, Linear, Quadratic, Sinusoidal. The sample size for each time point is set to 1,000 observations, so each embedding map is simulated a total of 7,776 times. All code required to reproduce the simulation results and the associated figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.

@@

The mean, median, standard deviation and interquartile range of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Tables \ref{table:simulation_results_mean} and \ref{table:simulation_results_median}. Because the PME, principal curve, and principal surface methods each attempt to estimate the manifold in question at each individual time point without allowing other time points to inform these estimates, they should result in similar deviations from the true underlying manifold, with differences in error resulting primarily from differing performances in fitting to the observed data. Meanwhile, because the LPME approach accounts for all time points simultaneously, this approach should result in lower mean squared distance values.
